\documentclass[a4paper,12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{enumitem}

% Title and Author Information
\title{Examination Element GRA4160 (30\%): \\ Mini Project and Presentation}
\date{}

\begin{document}

\maketitle

\noindent
You will work on a mini project in groups of 2-3 students.
The mini project will be a small-scale project where you will apply the methods you learn in the course to a real-world problem.

\bigskip
\noindent
To complete the project, you will need to perform the following steps:
\begin{itemize}
  \item Identify a problem or research question that you would like to address.
  \item Collect and preprocess data for your project.
  \item Choose a machine learning algorithm (or several) and implement it in your project.
  \item All the groups can book two meetings with me to discuss the project and get some feedback either in person or on Zoom during the semester.
  \item Make a presentation summarizing your findings and discussing the results.
  \item Present for the class (including an external examiner).
\end{itemize}

\noindent
The project is an important part of the course, as it allows you to demonstrate your understanding of machine learning concepts and techniques,
and to apply them in a practical setting.
It is also an opportunity for you to explore a topic of interest and to develop your skills as a data scientist.
The project will be graded and will count for 30\% of your final grade.

\vspace{1em}
\noindent\textbf{Note:} More information will be given during the course.

\end{document}
